Overrelaxed Sinkhorn--Knopp Algorithm for Regularized Optimal Transport

### 2017-10-20 paper.tex
small modifs

\section{Acceleration of local convergence rate}
\label{section:local}
In order to justify the acceleration of convergence that is observed in practice, we now study the local convergence rate of the overrelaxed algorithm, which follows from the classical convergence analysis of the linear SOR method. Our result involves the second largest eigenvalue of the matrix
\begin{equation}\label{eq:M1}
M_1= \diag(1\oslash \mu^1) \, \gamma^* \, \diag(1\oslash \mu^2)\, ^t\gamma^*
\end{equation}
where $\gamma^*$ is the solution to the regularized OT problem. We denote this second largest eigenvalue by $1-\eta$, which satisfies $\eta>0$ (see \cite{knight2008sinkhorn}).

@@

\begin{proof}
In this proof, we focus on the dual problem and we recall the relationship $\gamma^{\ell}=e^{\alpha^\ell/\epsilon}\gamma^0e^{\beta^\ell/\epsilon}$ between the iterates of the overrelaxed projection algorithm $\gamma^\ell$ and the  iterates $(\alpha^\ell,\beta^\ell)$ of the SOR algorithm on the dual problem~\eqref{SORdual}, initialized with $(\alpha^0,\beta^0)=(0,0)$. %First notice that if $\gamma^\ell \to \gamma^*$, then $\KL(\gamma^\ell,\gamma^0)\to\KL(\gamma^*,\gamma^0)$, which is equivalent to $E(\alpha^\ell,\beta^\ell)$ converges to the maximum of $E$.
The dual problem~\eqref{DROT} is invariant by translations of the form $(\alpha,\beta)\mapsto (\alpha-k,\beta+k)$, $k\in \IR$, but is otherwise strictly convex on any subspace which does not contain the line $\IR(\mathbf{1},-\mathbf{1})$ (this invariance cancels in the corresponding primal iterates). In order to deal with this invariance, consider the subspace $S$ of pairs of dual variables $(\alpha,\beta)$ that satisfy $\alpha_1=0$, let $\pi_S$ be the (non orthogonal) projection on $S$ of kernel $(\mathbf{1},-\mathbf{1})$ and let $(\alpha^*,\beta^*)\in S$ be a dual maximizer.

% 
Since the SOR iteration followed by $\pi_S$ is a smooth map, its local convergence properties are characterized by the local convergence of its linearization, which corresponds to the SOR method applied to the maximization of the quadratic Taylor expansion of the dual objective $E$ at $(\alpha^*,\beta^*)$. This defines an affine map $M_\theta : (\alpha^{\ell},\beta^{\ell})\mapsto (\alpha^{\ell+1},\beta^{\ell+1})$ whose spectral properties are well known~\cite{ciarlet1982introduction, young2014iterative} (see also \cite[chapter 4]{chizat2017these} for the specific case of convex minimization). For the case $\theta=1$, we obtain the matrix $M_1$ defined in \eqref{eq:M1}. Specifically, in the non strictly concave case~\cite{hadjidimos1985optimization}, we have that the operator $\pi_S\circ M_1^\ell$ converges at the linear rate $1-\eta$ towards the projector on $(\alpha^*,\beta^*)$ and that the convergence of $\pi_S\circ M_\theta^\ell$ is guaranteed for $\theta \in ]0,2[$, with the rate 
\[
f(\theta,\eta) = \begin{cases}
\theta-1 & \text{if $\theta>\theta^*$}\\
\frac12 \theta^2(1-\eta) - (\theta-1) +\frac12 \sqrt{(1-\eta)\theta^2(\theta^2(1-\eta)-4(\theta-1))} & \text{otherwise},
\end{cases}
\]
where $\theta^*\coloneqq2/(1+\sqrt{\eta})$ is the optimal parameter, for which $f(\theta^*,\eta)=(1-\sqrt{\eta})/(1+\sqrt{\eta})$. The function $f$ is plotted in Figure \ref{fig:eigtransform}.

@@

\begin{corollary}
The previous local convergence analysis applies to Algorithm \ref{thm:algo}, where the local convergence rate is governed by the choice of the target extrapolation parameter $\theta_0$.
\end{corollary}
\begin{proof}
What we need to show is that eventually one always has $\Theta(\gamma^\ell)=\theta_0$. This can be seen from the quadratic Taylor expansion $\varphi_{\theta_0}(1+z)=z^2(\theta_0 -\theta_0^2/2)+ o(z^2)$, which shows that for any choice of $\theta_0\in {]1,2[}$, there is a neighborhood of $1$ on which $\varphi_{\theta_0}(\cdot)$ is nonnegative.
\end{proof}

>>>>

\section{Acceleration of local convergence rate}
\label{section:local}
In order to justify the acceleration of convergence that is observed in practice, we now study the local convergence rate of the overrelaxed algorithm, which follows from the classical convergence analysis of the linear SOR method. Our result involves the second largest eigenvalue of the matrix
\begin{equation}\label{eq:M1}
M_1= \diag(1\oslash \mu^1) \, \gamma^* \, \diag(1\oslash \mu^2)\, ^t\gamma^*
\end{equation}
where $\gamma^*$ is the solution to the regularized OT problem (the largest eigenvalue is $1$, associated to the eigenvector $\mathbf{1}$). We denote this second largest eigenvalue by $1-\eta$, which satisfies $\eta>0$ (see \cite{knight2008sinkhorn}).

@@

\begin{proof}
In this proof, we focus on the dual problem and we recall the relationship $\gamma^{\ell}=e^{\alpha^\ell/\epsilon}\gamma^0e^{\beta^\ell/\epsilon}$ between the iterates of the overrelaxed projection algorithm $\gamma^\ell$ and the  iterates $(\alpha^\ell,\beta^\ell)$ of the SOR algorithm on the dual problem~\eqref{SORdual}, initialized with $(\alpha^0,\beta^0)=(0,0)$. %First notice that if $\gamma^\ell \to \gamma^*$, then $\KL(\gamma^\ell,\gamma^0)\to\KL(\gamma^*,\gamma^0)$, which is equivalent to $E(\alpha^\ell,\beta^\ell)$ converges to the maximum of $E$.
The dual problem~\eqref{DROT} is invariant by translations of the form $(\alpha,\beta)\mapsto (\alpha-k,\beta+k)$, $k\in \IR$, but is otherwise strictly convex on any subspace which does not contain the line $\IR(\mathbf{1},-\mathbf{1})$. In order to deal with this invariance (which cancels in the corresponding primal iterates), consider the subspace $S$ of pairs of dual variables $(\alpha,\beta)$ that satisfy $\alpha_1=0$, let $\pi_S$ be the (non orthogonal) projection on $S$ of kernel $(\mathbf{1},-\mathbf{1})$ and let $(\alpha^*,\beta^*)\in S$ be a dual maximizer.

% 
Since one SOR iteration is a smooth map, the local convergence properties of the SOR algorithm are characterized by the local convergence of its linearization, which here corresponds to the SOR method applied to the maximization of the quadratic Taylor expansion of the dual objective $E$ at $(\alpha^*,\beta^*)$. This defines an affine map $M_\theta : (\alpha^{\ell},\beta^{\ell})\mapsto (\alpha^{\ell+1},\beta^{\ell+1})$ whose spectral properties are well known~\cite{ciarlet1982introduction, young2014iterative} (see also \cite[chapter 4]{chizat2017these} for the specific case of convex minimization). For the case $\theta=1$, this corresponds to the matrix $M_1$ defined in \eqref{eq:M1}. Specifically, in the non strictly concave case~\cite{hadjidimos1985optimization}, we have that the operator $\pi_S\circ M_1^\ell$ converges at the linear rate $1-\eta$ towards the projector on $(\alpha^*,\beta^*)$ and that the convergence of $\pi_S\circ M_\theta^\ell$ is guaranteed for $\theta \in ]0,2[$, with the rate 
\[
f(\theta,\eta) = \begin{cases}
\theta-1 & \text{if $\theta>\theta^*$}\\
\frac12 \theta^2(1-\eta) - (\theta-1) +\frac12 \sqrt{(1-\eta)\theta^2(\theta^2(1-\eta)-4(\theta-1))} & \text{otherwise},
\end{cases}
\]
where $\theta^*\coloneqq2/(1+\sqrt{\eta})$ is the optimal parameter, for which $f(\theta^*,\eta)=(1-\sqrt{\eta})/(1+\sqrt{\eta})$. The function $f$ is plotted in Figure \ref{fig:eigtransform}.

@@

\begin{corollary}
The previous local convergence analysis applies to Algorithm \ref{thm:algo} and the local convergence rate is governed by the choice of the target extrapolation parameter $\theta_0$.
\end{corollary}
\begin{proof}
What we need to show is that eventually one always has $\Theta(\gamma^\ell)=\theta_0$. This can be seen from the quadratic Taylor expansion $\varphi_{\theta_0}(1+z)=z^2(\theta_0 -\theta_0^2/2)+ o(z^2)$, which shows that for any choice of $\theta_0\in {]1,2[}$, there is a neighborhood of $1$ on which $\varphi_{\theta_0}(\cdot)$ is nonnegative.
\end{proof}

%Of course, a difficulty is that the number $\eta$ is not known a priori. In the numerical experiments, we have noticed that
